---
title: "rdec - regression with damped exponential correlation"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{rdec - regression with damped exponential correlation}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

The model implemented here was described in Munoz A, Carey VJ, Schouten J,
Segal M, Rosner B, A parametric family of correlation structures for
the analysis of longitudinal data, Biometrics 48, 733-742, September 1992.
A JSTOR URL is [https://www.jstor.org/stable/2532340](https://www.jstor.org/stable/2532340).

We consider longitudinal responses $y_{ij}$, $j=1, \ldots n_i$, $i = 1, \ldots, N$.
The $n_i$-vector $y_i$ has elements that are assumed ordered in time with
timings recorded in $s_{ij}$.  A $p$-vector of predictors including an intercept
is provided for each observation in $x_{ij}$.

`rdec` fits the model 
\[
y_{ij} = x_{ij} \beta + e_{ij}
\]
in which the $n_i$-vectors $e_{i}$ are assumed multivariate Gaussian with
covariance $\sigma^2 R(\gamma, \theta; s_i)$.  For
$1 \leq g, h, \leq n_i$ the $g,h$ element of $R(\gamma, \theta; s_i)$
is given by
\[
Corr(y_{ig}, y_{ih}) = \gamma ^ { |s_{ig} - s_{ih} |^\theta }.
\]
Models with $\theta = 0$ are also called exchangeable or compound symmetry
models.  Models with $\theta = 1$ correspond to first-degree autoregressions (AR(1)
models).
As $\theta \rightarrow \infty$ the model approaches a first-order moving average.
`rdec` implements maximum likelihood estimation of all parameters $\beta$, $\sigma$,
$\gamma$, $\theta$.  Munoz et al in the paper referenced above provide empirical
evidence that members of the rdec model family can effectively approximate
second-order autoregressive (AR(2)) models and AR(1) models with random intercept.

# A simple simulation

## Building covariance matrices

The function `mycov` builds a correlation matrix
based on the rdec model, with observation times in `s`.
```{r dosim1}
library(MASS)
mycov = function(gam=.33, theta=.5, s=1:5) {
  d = as.matrix(dist(s, method="manhattan"))
  gam^(d^theta)
}
```

## Simulating and fitting a zero-mean model

The model can accommodate clusters of different sizes
and heterogeneous timings.  Here we simulate a zero-mean
model with fixed `gamma` and `theta` for common timings
for all elements of `N` clusters in `s`.
```{r lksim}
simgst_bal_zeromean = function(gam, theta, s1=1:5, N=100) {
  dat = mvrnorm(N, rep(0,length(s1)), mycov(gam=gam, theta=theta, s=s1))
  mdat = reshape2::melt(dat)
  ndat = dplyr::arrange(mdat, Var1)
  rdec::rdec(value~1, id=ndat$Var1, S=ndat$Var2, data=ndat)
}
```

Here is the result of a single simulation.
```{r do1}
set.seed(1234)
n1 = simgst_bal_zeromean(.33, .5)
summary(n1)
```

## Iterating the fitting process

Now we replicate the simulation process for 100 realizations.
```{r do2}
n2 = replicate(100, simgst_bal_zeromean(.33, .5))
```

A report on the first replicate is:
```{r do3}
rdec:::summary.rdec(n2[,1])
```

## Adequacy of standard error estimation

### Standard errors of $\gamma$ compared to sampling distribution

The distribution of estimates of $\gamma$ can be obtained using
```{r do4}
gams = apply(n2, 2, function(x) x$omega[1])
summary(gams)
```

The standard errors estimated by the procedure can be collected
as follows:
```{r do5}
getses = function(x) rdec:::summary.rdec(x)$se.omega
allses = apply(n2, 2, getses)
```

The standard errors should estimate
the standard deviation of estimates over the simulation
replicates.
```{r lk5}
sd(gams)
summary(allses[1,])
```

The median s.e. estimate for $\hat{\gamma}$ is close to the sampling s.d. for
the estimates of $\gamma$.  

### Similar process for $\theta$

```{r lkhes}
thes = apply(n2, 2, function(x) x$omega[2]) 
#vth = apply(n2,2,function(x)solve(-x$hessian)[4,4])
summary(allses[2,])
sd(thes)
```
